Activity monitoring data
============================================================
# 1- Load packages and setwd

``` {r}
library(knitr)
library(ggplot2)
library(data.table)
library(gridExtra)
setwd("C:/Users/camil/Desktop/DataScience/ReproducibleResearch-Week2")
```

# 2- Download file

```{r}
url<-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(url, "amd.zip")
```

First check the files that will be unzipped

```{r}
unzip("amd.zip", list=T)
```

Unzip the right file
```{r}
unzip("amd.zip")
```

# 3- Open and check table
```{r}
 amd<-read.csv("activity.csv")
```
 Check length and header names
 
```{r}
 dim(amd)
 colnames(amd)
```
# 4- Number of steps taken per day 

## 4-1- Total number of steps taken each day

```{r}
StepsDaySum<-tapply(amd$steps, amd$date, sum)
StepsDaySum<-as.data.frame(StepsDaySum)
colnames(StepsDaySum)<-c("steps")
StepsDaySum$date<-rownames(StepsDaySum)
rownames(StepsDaySum)<-seq_along(StepsDaySum$steps)
head(StepsDaySum)
```

## 4-2- Histogram of total number of steps taken each day
```{r}
StepsDaySum_hist<-hist(StepsDaySum$steps, main="Total number of steps taken each day", xlab="Number of steps")
StepsDaySum_hist
```

## 4-3- Mean and median of total numer of steps taken each day
```{r}
StepsDaySum_mean<-mean(StepsDaySum$steps, na.rm=T)
StepsDaySum_median<-median(StepsDaySum$steps, na.rm = T)
StepsDaySum_mean
StepsDaySum_median
```

# 5- Average daily activity pattern

## 5-1- Time series plot (5-min interval x average number of steps taken across all days)

First prepare the dataset
```{r}
StepsIntMean<-tapply(amd$steps, amd$interval, mean, na.rm=T)
StepsIntMean<-as.data.frame(StepsIntMean)
colnames(StepsIntMean)<-c("steps_mean")
StepsIntMean$interval<-rownames(StepsIntMean)
StepsIntMean$interval<-as.numeric(StepsIntMean$interval)
rownames(StepsIntMean)<-seq_along(StepsIntMean$interval)
head(StepsIntMean)
```
Then create the graph
```{r}
StepsIntMean_ggplot<-ggplot(data=StepsIntMean, aes(x=interval, y=steps_mean))+
    geom_point()+geom_line()+
    ggtitle("Mean number of steps per 5 minutes interval")+
    xlab("Time (minutes)")+ylab("Mean number of steps")+
    theme(panel.background = element_rect(fill = "white",colour = "white"),
          panel.grid.major = element_line(size = 0.2, linetype = 3,colour = "black"),
          panel.border = element_rect(fil=NA, colour="black"))
StepsIntMean_ggplot
```

## 5-2- Which 5-min interval, on average across all the days in the dataset, contains the maximum number of steps ? 
```{r}
StepsIntMean[StepsIntMean$steps_mean==max(StepsIntMean$steps_mean),]
```

# 6- Inputing missing values
## 6-1- Calculate and report total number of missing values in the dataset
```{r}
sum(is.na(amd$steps)==TRUE)
```

## 6-2- Strategy for filling in
First look where the NA are mostly represented -- this result is very long and not printed: 
```{r results="hide"}
is.na(amd$steps)
```
It looks like entire days are missing values, how many days is it ?
```{r}
sum(is.na(StepsDaySum$steps)==TRUE)
```
How many obersvations are there per day again ?
```{r}
length(unique(amd$interval))
```
Are all the missing informations coming from single days missing values ?
```{r}
sum(is.na(StepsDaySum$steps)==TRUE)*length(unique(amd$interval))
```

Yes. 
Two strategies are available to fill the NA values: fill each interval with either the mean or median value of the same time interval, to do that let's first see if the mean and median values are comparable.

We add a median column to the StepsIntMean dataset
```{r}
StepsIntMean$steps_median<-as.numeric(tapply(amd$steps, amd$interval, median, na.rm=T))
```

Let´s add a line to the graph: 
```{r}
StepsIntMean$steps_median<-as.numeric(tapply(amd$steps, amd$interval, median, na.rm=T))
StepsInt<-melt.data.table(as.data.table(StepsIntMean), id.vars=c("interval"))
    
StepsInt_ggplot<-ggplot(data=StepsInt)+
    geom_line(aes(x=interval,y=value,linetype=variable))+
    ggtitle("Mean and median number of steps per 5 minutes interval")+
    xlab("Time (minutes)")+ylab("Number of steps")+
    theme(panel.background = element_rect(fill = "white",colour = "white"),
          panel.grid.major = element_line(size = 0.2, linetype = 3,colour = "black"),
          panel.border = element_rect(fil=NA, colour="black"))
StepsInt_ggplot
```

The mean seems to be much higher than the median, and would probably skew the data towards hight values if the NA were filled in with the mean. However let´s make both datasets to see. 

## 6-3- Create a new dataset with filled in missing values

Fill in NAs with the median value of each interval and check if it worked
```{r}
amd2<-copy(amd)
for (i in amd2$interval) {
    amd2[amd2$interval==i & is.na(amd2$steps)==T, c("steps")]<-StepsIntMean[StepsIntMean$interval==i,c("steps_median")]
}
head(amd2)
```

Fill in NAs with the mean value of each interval and check if it worked
```{r}
amd3<-copy(amd)
for (i in amd3$interval) {
    amd3[amd3$interval==i & is.na(amd3$steps)==T, c("steps")]<-StepsIntMean[StepsIntMean$interval==i,c("steps_mean")]
}
head(amd3)
```

## 6-4- Histogram of total number of steps taken each day

Missing values filled with the median value of the same interval:
```{r}
StepsDaySum2<-tapply(amd2$steps, amd2$date, sum)
StepsDaySum2<-as.data.frame(StepsDaySum2)
colnames(StepsDaySum2)<-c("steps")
StepsDaySum2$date<-rownames(StepsDaySum2)
rownames(StepsDaySum2)<-seq_along(StepsDaySum2$steps)

StepsDaySum2_hist<-hist(StepsDaySum2$steps, main="Total number of steps taken each day \n(NA filled with median value of the same interval)", xlab="Number of steps")
StepsDaySum2_hist
```

Missing values filled with the mean value of the same interval:
```{r}
StepsDaySum3<-tapply(amd3$steps, amd3$date, sum)
StepsDaySum3<-as.data.frame(StepsDaySum3)
colnames(StepsDaySum3)<-c("steps")
StepsDaySum3$date<-rownames(StepsDaySum3)
rownames(StepsDaySum3)<-seq_along(StepsDaySum3$steps)

StepsDaySum3_hist<-hist(StepsDaySum3$steps, main="Total number of steps taken each day \n(NA filled with mean value of the same interval)",xlab="Number of steps")
StepsDaySum3_hist
```

To compare, histogram with missing values: 
```{r}
StepsdaySum_hist<-hist(StepsDaySum$steps, main="Total number of steps taken each day", xlab="Number of steps")
```

## 6-5- Mean and median of total number of steps taken each day

Missing values filled with the median value of the same interval:
```{r}
StepsDaySum2_mean<-mean(StepsDaySum2$steps, na.rm=T)
StepsDaySum2_median<-median(StepsDaySum2$steps, na.rm = T)
StepsDaySum2_mean
StepsDaySum2_median
```

Missing values filled with the median value of the same interval:
```{r}
StepsDaySum3_mean<-mean(StepsDaySum3$steps, na.rm=T)
StepsDaySum3_median<-median(StepsDaySum3$steps, na.rm = T)
StepsDaySum3_mean
StepsDaySum3_median
```

To compare, mean and median with missing values: 
```{r}
StepsDaySum_mean
StepsDaySum_median
```

### 6-5-1- Do these values differ from the first estimates ?

If the missing values are filled in with the median value of each interval, they differ by about 500 steps, so approximately 5% of the total number of steps. 
If the missing values are filled in with the mean value of each interval, the mean value doesn´t change, which is awaited and the median is also almost the same.

### 6-5-2- What is the impact of imputing missing data on the estimates of the total daily number of steps ? 

If one inputs missing data from the mean value of the same time interval, one probably slightly overestimates the total number of steps taken, however the overestimation is not that significant when considered over the whole day. If one considers times at which a relatively low numbers of steps are taken, such as for example in the hours after midnight, then the results are strongly skewed when filled in by the mean, as the mean is falsely raised by outliers.

# 7-Differences in activity patterns between weekdays and weekends

## 7-1- New factor variable 

The extra "day" variable was created as a control. 
```{r}
amd$day<-weekdays(as.Date(amd$date))
amd$we<-weekdays(as.Date(amd$date))
weekdays<-c("Monday","Tuesday","Wednesday","Thursday","Friday")
amd[is.element(amd$we, weekdays)==F,c("we")]<-c("weekend")
amd[is.element(amd$we, weekdays)==T,c("we")]<-c("weekday")
```

## 7-2- Time series plot

I chose to separate my dataframe into two separate data frames, make the graphics and arrange them together afterwards.

First the dataframe and timeseries plot for weekdays: 
```{r}
amdwd<-amd[is.element(amd$day, weekdays)==T,]

StepsIntWD<-tapply(amdwd$steps, amdwd$interval, mean, na.rm=T)
StepsIntWD<-as.data.frame(StepsIntWD)
colnames(StepsIntWD)<-c("steps_mean")
StepsIntWD$interval<-rownames(StepsIntWD)
StepsIntWD$interval<-as.numeric(StepsIntWD$interval)
rownames(StepsIntWD)<-seq_along(StepsIntWD$interval)

StepsIntWD_ggplot<-ggplot(data=StepsIntWD, aes(x=interval, y=steps_mean))+
    geom_point()+geom_line()+
    ggtitle("Mean numer of steps per 5 minutes interval on weekdays")+
    xlab("Time (minutes)")+ylab("Mean number of steps")+
    theme(panel.background = element_rect(fill = "white",colour = "white"),
          panel.grid.major = element_line(size = 0.2, linetype = 3,colour = "black"),
          panel.border = element_rect(fil=NA, colour="black"))
```

Then the dataframe and timeseries plot for weekends:
```{r}
amdwe<-amd[is.element(amd$day, weekdays)==F,]                                  

StepsIntWE<-tapply(amdwe$steps, amdwe$interval, mean, na.rm=T)
StepsIntWE<-as.data.frame(StepsIntWE)
colnames(StepsIntWE)<-c("steps_mean")
StepsIntWE$interval<-rownames(StepsIntWE)
StepsIntWE$interval<-as.numeric(StepsIntWE$interval)
rownames(StepsIntWE)<-seq_along(StepsIntWE$interval)

StepsIntWE_ggplot<-ggplot(data=StepsIntWE, aes(x=interval, y=steps_mean))+
    geom_point()+geom_line()+
    ggtitle("Mean numer of steps per 5 minutes interval on weekends")+
    xlab("Time (minutes)")+ylab("Mean number of steps")+
    theme(panel.background = element_rect(fill = "white",colour = "white"),
          panel.grid.major = element_line(size = 0.2, linetype = 3,colour = "black"),
          panel.border = element_rect(fil=NA, colour="black"))
```

Then merge the graphis: 
```{r}
StepsIntMerge_ggplot<-grid.arrange(StepsIntWD_ggplot, StepsIntWE_ggplot)
```
